Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Before moving case study 1 and 2 to a separate document

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters, and $p=3$ columns. Column $X1$ is completely observed, column $X2$ is systematically missing in cluster 2, and column $X3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method or discard the incomplete units altogether (i.e., list-wise deletion/complete case analysis). Further, with the systematic missingness in $X2$, it is impossible to fit a multilevel model since we cannot estimate the intercept of cluster 2. We would have to exclude this cluster from our analyses entirely to obtain any results. Obviously, excluding observations is not a desirable workflow.

@@

Ignoring the missingness in analyses can be extremely harmful to inferences. Complete case analysis can introduce bias in statistical inferences and lowers statistical power. Instead, the missingness should be accommodated \emph{before} or _within_ the analysis of scientific interest. Especially the former is very generic and popular. [VJ: add why multiple values are imputed.] Imputing (i.e., filling in) the missing values separates the missing data problem from the scientific problem: missing data are replaced by plausible values whereafter the completed data is analyzed as if it were completely observed. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.

@@

---

\begin{center}
REVIEW UNTIL THIS POINT SVP
\end{center}

---

>>>>

The process of analyzing multilevel data is further complicated when not all data entries are observed. Just as with single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for clustered missingness. Therefore, incomplete multilevel data can be categorized into two general patterns: systematic missingness and sporadic missingness [@resc13]. Systematic missingness implies that one or more variables are never observed in a certain cluster. With sporadic missingness there may be observed data for some but not all units in a cluster [@buur18; @jola18]. We have visualized this difference in Figure 1, which shows an $n \times p$ set $\mathbf{X} = X_1, \dots, X_p$, with $n$ units distributed over $N$ clusters and $p$ variables. Column $X_1$ is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.

<!-- Obviously, excluding observations is not a desirable workflow.  -->

@@

Since excluding observations is not a desirable workflow, the missingness in multilevel data should be accommodated \emph{before} or _within_ the analysis of scientific interest. In this paper, we focus on the former approach: imputing (i.e., filling in) the missing data with plausible values, whereafter the completed data may be analyzed as if it were completely observed. Imputation separates the missing data problem from the scientific problem, which makes the missing data strategy very generic and popular. If each missing value is replaced multiple times, the resulting inferences may validly convey the uncertainty due to missingness [c.f. @rubi76]. The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. In this paper, we will discuss how to use \pkg{mice} in the context of multilevel data.

@@

<!-- --- -->

<!-- \begin{center} -->
<!-- REVIEW UNTIL THIS POINT SVP -->
<!-- \end{center} -->

<!-- --- -->

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Before moving case study 1 and 2 to a separate document

\begin{table}[tb]
\caption{Key concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Therefore,
incomplete multilevel data can be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}.
Systematic missingness implies that one or more variables are never
observed in a certain cluster. With sporadic missingness there may be
observed data for some but not all units in a cluster
\citep{buur18, jola18}. We have visualized this difference in Figure 1,
which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
\(n\) units distributed over \(N\) clusters, and \(p=3\) columns. Column
\(X1\) is completely observed, column \(X2\) is systematically missing
in cluster 2, and column \(X3\) is sporadically missing. To analyze
these incomplete data, we have to take the nature of the missingness and
the cluster structure into account. For example, the sporadic
missingness in \(X3\) could be easily amended if this would be a
cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method or discard the incomplete units
altogether (i.e., list-wise deletion/complete case analysis). Further,
with the systematic missingness in \(X2\), it is impossible to fit a
multilevel model since we cannot estimate the intercept of cluster 2. We
would have to exclude this cluster from our analyses entirely to obtain
any results. Obviously, excluding observations is not a desirable
workflow.

@@

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis can introduce bias in statistical
inferences and lowers statistical power. Instead, the missingness should
be accommodated \emph{before} or \emph{within} the analysis of
scientific interest. Especially the former is very generic and popular.
{[}VJ: add why multiple values are imputed.{]} Imputing (i.e., filling
in) the missing values separates the missing data problem from the
scientific problem: missing data are replaced by plausible values
whereafter the completed data is analyzed as if it were completely
observed. The \proglang{R} package \pkg{mice} has become the de-facto
standard for imputation by chained equations, which iteratively solves
the missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
the context of multilevel data.

>>>>

\begin{table}[tb]
\caption{Concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Therefore,
incomplete multilevel data can be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}.
Systematic missingness implies that one or more variables are never
observed in a certain cluster. With sporadic missingness there may be
observed data for some but not all units in a cluster
\citep{buur18, jola18}. We have visualized this difference in Figure 1,
which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
\(n\) units distributed over \(N\) clusters and \(p\) variables. Column
\(X_1\) is completely observed, column \(X_2\) is systematically missing
in cluster 2, and column \(X_3\) is sporadically missing. To analyze
these incomplete data, we have to take the nature of the missingness and
the cluster structure into account. For example, the sporadic
missingness in \(X_3\) could be easily amended if this would be a
cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X_3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method, or discard the incomplete units
altogether (i.e., complete case analysis). Complete case analysis can
however introduce bias in statistical inferences and lowers statistical
power. Further, with the systematic missingness in \(X_2\), it would be
impossible to fit a multilevel model without accommodating the
missingness in some way. Complete case analysis in that case would mean
excluding the entire cluster from the analyses. The wrong choice of
missing data handling method can thus be extremely harmful to the
inferences.

@@

Since excluding observations is not a desirable workflow, the
missingness in multilevel data should be accommodated \emph{before} or
\emph{within} the analysis of scientific interest. In this paper, we
focus on the former approach: imputing (i.e., filling in) the missing
data with plausible values, whereafter the completed data may be
analyzed as if it were completely observed. Imputation separates the
missing data problem from the scientific problem, which makes the
missing data strategy very generic and popular. If each missing value is
replaced multiple times, the resulting inferences may validly convey the
uncertainty due to missingness \citep[c.f.][]{rubi87}. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
the context of multilevel data.
